---
title: "Your Awesome Title"
author: "Nathaniel Haulk^1^"
date: '`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    includes:
      before_body: template/doc_prefix.tex
      in_header: template/preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2:
    toc: no
    reference_docx: template/nature_style.docx
    number_sections: no
fontsize: 12pt
linestretch: 1.5
link-citations: yes
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ecology.csl
bibliography: ref.bib
always_allow_html: yes
links-as-notes: true
notes-after-punctuation: false
---

```{r knitr_options, echo=FALSE}
library(knitr)
# rstudio will set the folder where .Rmd file seats as work directory
# set it back to the folder where .Rproj seats
opts_knit$set(root.dir = normalizePath("../")) 
opts_chunk$set(fig.align = 'center', cache = FALSE, warning = FALSE,
  message = TRUE, echo = FALSE, fig.pos = "H", out.extra = "")
options(digits = 3, width = 88,
        # use pdf file automatically when compile to pdf
        knitr.graphics.auto_pdf = TRUE, 
        knitr.kable.NA = '')

# download template files if not available
tpl_1 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/Doc/template/preamble.tex'
tpl_2 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/Doc/template/doc_prefix.tex'
tpl_3 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/Doc/template/nature_style.docx'
# bib_1 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/template/ref.bib'
# change directory accordingly
if(!file.exists(tpl_1f <- 'template/preamble.tex')) download.file(tpl_1, tpl_1f)
if(!file.exists(tpl_2f <- 'template/doc_prefix.tex')) download.file(tpl_2, tpl_2f)
if(!file.exists(tpl_3f <- 'template/nature_style.docx')) download.file(tpl_3, tpl_3f)

if(knitr::is_latex_output() | knitr::is_html_output()){
  library(kableExtra)
} else {
  options(kableExtra.auto_format = FALSE) # for docx
}
```

\footnotesize 

^1^Department of Biological Sciences, Louisiana State University, Baton Rouge, LA, USA  

email: nhaulk1@lsu.edu; 232 Life Science Building, Baton Rouge, LA 70803

\normalsize

**Running headline**: Environment and species richness

**Abstract**: Your awesome abstract here.

\clearpage

# Introduction

The first case of Covid-19 was recorded in the United States in January 22, 2020 according to the CDC. As of December 5, 2021, the United States has reported more than 780000 deaths due to Covid-19, more than any other country [@ny_times_coronavirus_2021]. Initially, non-pharmaceutical methods were used to help stop the spread of Covid-19, including the closing of many public spaces, encouraging social distancing, and encouraging the use of masks to help reduce transmission rates. Through time, social distancing and mask mandate policies changed as Covid-19 became the new norm, leading to patterns of increasing and decreasing Covid-19 rates in the United States [@rebeiro_impact_2021]. The current widespread availability of vaccines helping to fight Covid-19 has also helped to reduce the overall transmission and resulting mortality [@haas_infections_2021]. Transmission and Covid-19 rates have increased post vaccination due to mutations in spike proteins that have allowed for new variants of COVID-19 to spread [@sheikh_sars-cov-2_2021, @monajjemi_delta_nodate]. As of December 5th, 2021, the Delta variant, is the most common strain of COVID-19, representing more than 99 percent of cases [@cdc_covid_2020].

Mask mandates, social distancing, and vaccination have all been proven to reduce the transmission of Covid-19 [@nguyen_mask_2021;@lyu_community_2020;@ferguson_report_2020;@haas_infections_2021]. Areas that adopted the use of masks early tended to have lower overall COVID-19 rates compared to those that did not. Given the efficacy of masks  preventing transmission as well as the high-availability to everyone, masks have been one of the best tools in fighting the spread of Covid-19 [@rebeiro_impact_2021]. Despite the clear benefits, masks have been met with much opposition across the United States, especially in politically conservative leaning counties [@he_why_2021; @kahane_politicizing_2021]. These counties tend to be more rural and more against mask mandates than their urban counterparts [@pro_us_2021].

Spatial Heterogeneity, or the uneven distribution of populations across space, is important for predicting disease dynamics in pathogens like Covid-19. Areas with low spatial heterogeneity tend to be better environments for a disease to persist when compared to high spatial heterogeneous environments [@hagenaars_spatial_2004]. Populations with higher densities tend to have higher disease spread assuming the disease transmission is not based on frequency of contact [@scott_impact_1988]. Density dependent disease like Covid-19 should therefore have a harder time spreading in rural areas with higher heterogeneity and lower population densities. However, with resistance to preventative measures such as masks and vaccines, cases may be higher in rural areas than their urban counterparts [@sun_rural-urban_2021].

With changes in population density and spacial spread, the basic reproductive number, $R_0$, should change as a result. The basic reproductive number represents the number of predicted infections that can result from one infection [@delamater_complexity_2019]. While studies have already looked into how $R_0$ changes by county, no recent studies have looked at the effective reproductive number on a county level [@ives_estimating_2021]. More importantly, not much is yet known about the new Omicron variant first introduced into the US on December 1, 2021. The Omicron variant is predicted to be more infectious than previous variants of Covid-19 and is predicted to be more resistant to the vaccines we currently have available [@cdc_omicron_2021]. The Reproductive number of the Omicron variant is not calculable at such a low prevalence due to the lack of data.However, predictions on how the new variant will spread provides vital information on preventing new infections and reducing overall mortality. 

While data is presented on county-level Covid rates, few studies have taken this further to investigate the how the infection rates in urban and rural areas compare. With the Omicron variant just recently emerging, no studies have yet to be published examining the predicted rate of spread. In this study, I seek to investigate the total infection prevalence per county. I also compare the prevalence per county in rural and urban areas, as well as those with and without mask mandates. Finally, I seek to use previous predictions of R0 to calculate Rt in its current state. Using these current Rt values, I then infer predicted Rt values for Omicron for each county to highlight at-risk counties. 


# Methods

Here is the method section. You can include equations easily. For inline equations, use $\text{var}(X) = p(1-p)$. For display equation, use

$$\text{var}(X) = p(1-p)$$

## Results


#### Tables

Insert tables by `kable` in knitr package in R. Then cross-reference it back with: see Table \@ref(tab:tableName).

(ref:captionTable1) **Model coefficients of leaf senescence based on in situ data**. No space at the end of this line!!

```{r tableName,results='asis',echo=FALSE, message=F, warning=F}

```

Put results inline, e.g. the mean species richness is.

#### Figures

Insert figure by code chunk. And cross-ref it back as Figure \@ref(fig:figName).

(ref:captionFig1) **No - or _ in the caption ref.** No space at the end of this line!

```{r figName, fig.width=7, fig.asp=1, fig.cap="(ref:captionFig1)"}

```


Or if you already have the figure: 

And cite it as Figure \@ref(fig:fig2).

(ref:captionFig2) **No - or _ in the caption ref.** No space at the end of this line!

```{r fig2, out.width='70%', fig.cap="(ref:captionFig2)"}

```

More details can be found at [here](https://bookdown.org/yihui/bookdown/).

# References

<div id="refs"></div>

\clearpage

\setcounter{page}{0}
\pagenumbering{arabic}
\setcounter{page}{1}


\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand {\thetable}{S\arabic{table}}
\renewcommand {\thefigure}{S\arabic{figure}}


# Supporting Information

## Figures

(ref:captionFigS1) **No - or _ in the caption ref.** No space at the end of this line!

```{r figS1, out.width='70%', fig.cap="(ref:captionFigS1)"}

```


