---
title: "County Level Predictors for Covid-19 Prevelance During the Introduction of Omnicron"
author: "Nathaniel Haulk^1^"
date: '`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    includes:
      before_body: template/doc_prefix.tex
      in_header: template/preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2:
    toc: no
    reference_docx: template/nature_style.docx
    number_sections: no
fontsize: 12pt
linestretch: 1.5
link-citations: yes
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ecology.csl
bibliography: ref.bib
always_allow_html: yes
links-as-notes: true
notes-after-punctuation: false
---

```{r knitr_options, echo=FALSE}
library(knitr)
# rstudio will set the folder where .Rmd file seats as work directory
# set it back to the folder where .Rproj seats
opts_knit$set(root.dir = normalizePath("../")) 
opts_chunk$set(fig.align = 'center', cache = FALSE, warning = FALSE,
  message = TRUE, echo = FALSE, fig.pos = "H", out.extra = "")
options(digits = 3, width = 88,
        # use pdf file automatically when compile to pdf
        knitr.graphics.auto_pdf = TRUE, 
        knitr.kable.NA = '')

# download template files if not available
tpl_1 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/Doc/template/preamble.tex'
tpl_2 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/Doc/template/doc_prefix.tex'
tpl_3 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/Doc/template/nature_style.docx'
# bib_1 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/template/ref.bib'
# change directory accordingly
if(!file.exists(tpl_1f <- 'template/preamble.tex')) download.file(tpl_1, tpl_1f)
if(!file.exists(tpl_2f <- 'template/doc_prefix.tex')) download.file(tpl_2, tpl_2f)
if(!file.exists(tpl_3f <- 'template/nature_style.docx')) download.file(tpl_3, tpl_3f)

if(knitr::is_latex_output() | knitr::is_html_output()){
  library(kableExtra)
} else {
  options(kableExtra.auto_format = FALSE) # for docx
}
```

\footnotesize 

^1^Department of Biological Sciences, Louisiana State University, Baton Rouge, LA, USA  

\* **Corresponding author**, email: nhaulk1@lsu.edu; 232 Life Science Building, Baton Rouge, LA 70803

\normalsize

**Running headline**: Environment and species richness

**Abstract**: Your awesome abstract here.

\clearpage

# Introduction

The first case of Covid-19 was recorded in the United States in January 22, 2020 according to the CDC. As of December 5, 2021, the United States has reported more than 780000 deaths due to Covid-19, more than any other country [@new_york_times_coronavirus_2021]. Initially, non-pharmaceutical methods were used to help stop the spread of Covid-19, including the closing of many public spaces, encouraging social distancing, and encouraging the use of masks to help reduce transmission rates. Through time, social distancing and mask mandate policies changed as Covid-19 became the new norm, leading to patterns of increasing and decreasing Covid-19 rates in the United States [@rebeiro_impact_2021]. The current widespread availability of vaccines helping to fight Covid-19 has also helped to reduce the overall transmission and resulting mortality [@haas_infections_2021]. Transmission and Covid-19 rates have increased post vaccination due to mutations in spike proteins that have allowed for new variants of COVID-19 to spread [@sheikh_sars-cov-2_2021, @monajjemi_delta_nodate]. As of December 5th, 2021, the Delta variant, is the most common strain of COVID-19, representing more than 99 percent of cases [@cdc_covid_2020].

Mask mandates, social distancing, and vaccination have all been proven to reduce the transmission of Covid-19 [@nguyen_mask_2021;@lyu_community_2020;@ferguson_report_2020;@haas_infections_2021]. Areas that adopted the use of masks early tended to have lower overall COVID-19 rates compared to those that did not. Given the efficacy of masks  preventing transmission as well as the high-availability to everyone, masks have been one of the best tools in fighting the spread of Covid-19 [@rebeiro_impact_2021]. Despite the clear benefits, masks have been met with much opposition across the United States, especially in politically conservative leaning counties [@he_why_2021; @kahane_politicizing_2021]. These counties tend to be more rural and more against mask mandates than their urban counterparts [@pro_us_2021].

Spatial Heterogeneity, or the uneven distribution of populations across space, is important for predicting disease dynamics in pathogens like Covid-19. Areas with low spatial heterogeneity tend to be better environments for a disease to persist when compared to high spatial heterogeneous environments [@hagenaars_spatial_2004]. Populations with higher densities tend to have higher disease spread assuming the disease transmission is not based on frequency of contact [@scott_impact_1988]. Density dependent disease like Covid-19 should therefore have a harder time spreading in rural areas with higher heterogeneity and lower population densities. However, with resistance to preventative measures such as masks and vaccines, cases may be higher in rural areas than their urban counterparts [@sun_rural-urban_2021].

With changes in population density and spacial spread, the basic reproductive number, $R_0$, should change as a result. The basic reproductive number represents the number of predicted infections that can result from one infection [@delamater_complexity_2019]. While studies have already looked into how $R_0$ changes by county, no recent studies have looked at the effective reproductive number on a county level [@ives_estimating_2021]. More importantly, not much is yet known about the new Omicron variant first introduced into the US on December 1, 2021. The Omicron variant is predicted to be more infectious than previous variants of Covid-19 and is predicted to be more resistant to the vaccines we currently have available [@cdc_omicron_2021]. The Reproductive number of the Omicron variant is not calculable at such a low prevalence due to the lack of data. However, predictions on how the new variant will spread provides vital information on preventing new infections and reducing overall mortality. 

While data is presented on county-level Covid-19 rates, few studies have taken this further to investigate the how the infection rates in urban and rural areas compare. With the Omicron variant just recently emerging, no studies have yet to be published examining the predicted rate of spread. In this study, I seek to investigate the total infection prevalence per county. I also compare the prevalence per county in rural and urban areas, as well as those with and without mask mandates. Finally, I seek to use previous predictions of R0 to calculate Rt in its current state. Using these current Rt values, I then infer predicted Rt values for Omicron for each county to highlight at-risk counties. 


# Methods

All code and graphs were written using the programming language R and maps of the United States were modeled using the package USmaps[@r_core_team_r_2021;@lorenzo_usmap_2021]. For data on Covid-19 cases by county, I used the time series data provided by the New York Times [@new_york_times_coronavirus_2021]. For total population estimates, I used US Census data predictions for 2021 [@united_states_census_bureau_county_2021]. Total cases were mapped out as the number of cases in each county divided by the total population to get the total percentage of infections (Figure \@ref(fig:fig1)). Rolling averages were mapped out based on reported cases within the last 30 days (Figure \@ref(fig:fig2)). Mask mandate data used was collected in July of 2020[@wright_tracking_2020]. Data from the early infection was chosen and graphed to show how early responses to Covid-19 have led difference in the total percentage of cases. A t-test was performed to test the significance between the two groups and bar graph was made for visualization (Figure \@ref(fig:figS1)).  Rural data was graphed using two criteria, based on HRSA classifications and the census Bureaus classifications of urban and rural areas [@hrsa_defining_2021;@united_states_census_bureau_2010_2021]. Both were graphed as a map of the US to show county classification as well as a bar graph to compare the average total case percentage (Figure \@ref(fig:figS2) and \@ref(fig:figS3)). An ANOVA test was performed on the data classified based on the Census Bureau Data distinction of urban and rural to test for significance. A t-test was performed on the data classified based on the HRSA distinction of urban and rural to test for significance. R Vaccine data was collected from the CDC to calculate the proportion of susceptible individuals [@cdc_covid-19_2021].

The effective reproductive number ($R_t$) was calculated using basic reproductive values($R_0$) for each county predicted in a previous studies (Figure \@ref(fig:fig3)) [@ives_estimating_2021] . Using previous models and calculations of $R_0$, the formula
$$R_0 = \frac{\beta*\lambda}{(\gamma+\mu+\epsilon+\sigma)*(\tau+\mu)}$$
was used to calculate the $\beta$ value, or the transmission rate, per county (see Table \@ref(tab:suptable1) for parameter descriptions) [@ahmed_mathematical_2021]. Beta values were than used to calculate using the formula 
$$R_t = (1-p)*\frac{S}{N}*\frac{\beta*\lambda}{(\gamma+\mu+\epsilon+\sigma)*(\tau+\mu)}$$
where S represents individuals Suscpetible to the disease and N represents the total population size in that count. Parameter p represents the proportion of vaccinated individuals per county [@cdc_covid-19_2021]. The effective reproductive number for the Omicron was then calculated using the same formula, with higher overall transmission rates and lower mask mandate usage to reflect the increase in R0 seen when the Delta variant of Covid-19 was first introduced into the United States(Figure \@ref(fig:fig4)) [@liu_reproductive_2021].

```{r sourcecode, echo = FALSE, results='hide', message=FALSE}
source("Code/Population_Data.R")

source("Code/County_R0.R")
```

# Results

#### Figures

Insert figure by code chunk. And cross-ref it back as Figure \@ref(fig:fig1).

(ref:captionfig1) **No - or _ in the caption ref.** No space at the end of this line!

```{r fig1, fig.width=7, fig.asp=1, fig.cap="(ref:captionfig1)"}
print(us_totinf)
```

(ref:captionfig2) **No - or _ in the caption ref.** No space at the end of this line!

```{r fig2, , out.width='70%', fig.cap="(ref:captionfig2)"}
print(us_curinf)
```

(ref:captionfig3) **No - or _ in the caption ref.** No space at the end of this line!

```{r fig3, fig.width=7, fig.asp=1, fig.cap="(ref:captionfig3)"}
print(map_rt_cur)
```

(ref:captionfig4) **No - or _ in the caption ref.** No space at the end of this line!

```{r fig4, fig.width=7, fig.asp=1, fig.cap="(ref:captionfig4)"}
print(map_rt_omeg)
```

# Discussion

I used current data on Covid-19, effective

Focus on these counties

Masks do work.

Vaccine data all over the place. 

Potential error

# Conclusion

# References

<div id="refs"></div>

\clearpage

\setcounter{page}{0}
\pagenumbering{arabic}
\setcounter{page}{1}


\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand {\thetable}{S\arabic{table}}
\renewcommand {\thefigure}{S\arabic{figure}}

# Supporting Information

## Tables 


Table: (#tab:suptable1) Parameter descriptions and values used to calculate $R_0$ [@ahmed_mathematical_2021].

Parameter             Value               Description
------------         -------------     ------------------ 
$\beta$                 Varies          Transmission rate 
$\lambda$               $.0025$           Births into system
$\gamma$                $2.01*10^{-4}$  Transfer from the exposed class to the quarantined class
$\mu$                   $.0015$           Natural Mortality Rate
$\epsilon$              $.45$             Transfer from the exposed class to the symptomatic infected class
$\sigma$                $.067$            Transfer from the exposed class to the asymptomatic infected class
$\tau$                  $2*10^{-4}$     Transfer from the susceptible class to the quarantine class
p                       Varies          Proportion vaccinated


(ref:captionFigS1) **No - or _ in the caption ref.** No space at the end of this line!

## Figures

```{r figS1, fig.width=7, fig.asp=1, fig.cap="(ref:captionFigS1)", message=FALSE}
library('gridExtra')
grid.arrange(us_mask,bar_mask,nrow =2)
```

(ref:captionFigS2) **No - or _ in the caption ref.** No space at the end of this line!

```{r figS2, fig.width=7, fig.asp=1, fig.cap="(ref:captionFigS2)"}

grid.arrange(US_RU_HRSA, Bar_RU_HRSA ,nrow =2)

```

(ref:captionFigS3) **No - or _ in the caption ref.** No space at the end of this line!

```{r figS3, fig.width=7, fig.asp=1, fig.cap="(ref:captionFigS3)"}
grid.arrange(US_RU_Cen,Bar_RU_Cen,nrow =2)
```
