---
title: "Your Awesome Title"
author: "Nathaniel Haulk^1^"
date: '`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    includes:
      before_body: template/doc_prefix.tex
      in_header: template/preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2:
    toc: no
    reference_docx: template/nature_style.docx
    number_sections: no
fontsize: 12pt
linestretch: 1.5
link-citations: yes
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ecology.csl
bibliography: ref.bib
always_allow_html: yes
links-as-notes: true
notes-after-punctuation: false
---

```{r knitr_options, echo=FALSE}
library(knitr)
# rstudio will set the folder where .Rmd file seats as work directory
# set it back to the folder where .Rproj seats
opts_knit$set(root.dir = normalizePath("../")) 
opts_chunk$set(fig.align = 'center', cache = FALSE, warning = FALSE,
  message = TRUE, echo = FALSE, fig.pos = "H", out.extra = "")
options(digits = 3, width = 88,
        # use pdf file automatically when compile to pdf
        knitr.graphics.auto_pdf = TRUE, 
        knitr.kable.NA = '')

# download template files if not available
tpl_1 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/Doc/template/preamble.tex'
tpl_2 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/Doc/template/doc_prefix.tex'
tpl_3 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/Doc/template/nature_style.docx'
# bib_1 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/template/ref.bib'
# change directory accordingly
if(!file.exists(tpl_1f <- 'template/preamble.tex')) download.file(tpl_1, tpl_1f)
if(!file.exists(tpl_2f <- 'template/doc_prefix.tex')) download.file(tpl_2, tpl_2f)
if(!file.exists(tpl_3f <- 'template/nature_style.docx')) download.file(tpl_3, tpl_3f)

if(knitr::is_latex_output() | knitr::is_html_output()){
  library(kableExtra)
} else {
  options(kableExtra.auto_format = FALSE) # for docx
}
```

\footnotesize 

^1^Department of Biological Sciences, Louisiana State University, Baton Rouge, LA, USA  

email: nhaulk1@lsu.edu; 232 Life Science Building, Baton Rouge, LA 70803

\normalsize

**Running headline**: Environment and species richness

**Abstract**: Your awesome abstract here.

\clearpage

# Introduction

The first case of Covid-19 was recorded in the United States in January 22, 2020 according to the CDC. As of December 5, 2021, the United States has reported more than 780000 deaths due to Covid-19, more than any other country [@nytimes_coronavirus_2021]. Initially, non-pharmaceutical methods were used to help stop the spread of Covid-19, including the closing of many public spaces, encouraging social distancing, and encouraging the use of masks to help reduce transmission rates. However

Mask mandates and social distancing have both been proven to reduce the transmission of Covid-19 [@nguyen_mask_2021;@lyu_community_2020;@ferguson_report_2020]

Omnicron

In this study, I investigate the total infection prevalence per county. I also compare the prevalence per county 


# Methods

Here is the method section. You can include equations easily. For inline equations, use $\text{var}(X) = p(1-p)$. For display equation, use

$$\text{var}(X) = p(1-p)$$

## Results


#### Tables

Insert tables by `kable` in knitr package in R. Then cross-reference it back with: see Table \@ref(tab:tableName).

(ref:captionTable1) **Model coefficients of leaf senescence based on in situ data**. No space at the end of this line!!

```{r tableName,results='asis',echo=FALSE, message=F, warning=F}

```

Put results inline, e.g. the mean species richness is.

#### Figures

Insert figure by code chunk. And cross-ref it back as Figure \@ref(fig:figName).

(ref:captionFig1) **No - or _ in the caption ref.** No space at the end of this line!

```{r figName, fig.width=7, fig.asp=1, fig.cap="(ref:captionFig1)"}

```


Or if you already have the figure: 

And cite it as Figure \@ref(fig:fig2).

(ref:captionFig2) **No - or _ in the caption ref.** No space at the end of this line!

```{r fig2, out.width='70%', fig.cap="(ref:captionFig2)"}

```

More details can be found at [here](https://bookdown.org/yihui/bookdown/).

# References

<div id="refs"></div>

\clearpage

\setcounter{page}{0}
\pagenumbering{arabic}
\setcounter{page}{1}


\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand {\thetable}{S\arabic{table}}
\renewcommand {\thefigure}{S\arabic{figure}}


# Supporting Information

## Figures

(ref:captionFigS1) **No - or _ in the caption ref.** No space at the end of this line!

```{r figS1, out.width='70%', fig.cap="(ref:captionFigS1)"}

```


