---
title: "Homework 6"
author: "Nathaniel Haulk"
date: "11/8/2021"
output: pdf_document
---
# Question 1

## Part a
```{r 1A}
## Data from the website
x = c(110.5, 105.4, 118.1, 104.5, 93.6, 84.1, 77.8, 75.6)
y = c(5.755, 5.939, 6.010, 6.545, 6.730, 6.750, 6.899, 7.862)

## binds x and y into a dataframe
d = data.frame(c(x,y))

## Creates variable fit_d that shows the slope and the y-intercept
fit_d = lm(y~x, data = d)

print(fit_d)
```
The least squares estimate is of $\beta_1$ is -.037. This value represent the best fit of the trend of the data that reduces the distance from all of the points to the line itself

## Part B
```{r 1B}
## Runs an ANOVA test
anovad = anova(fit_d)

print(anovad)

## used to get the t-test p-value
summaryd = summary(fit_d)

print(summaryd)
```

Both the F-test and the T-test show a P-value <. 01 confirming the alternative hypothesis. This means that $H_a\neq0$ is true. 

## Part C

```{r Q1C}
## Calculates the t_(n-2,a/2) value
print(qt(.05/2, 8-2))

## Calculate the confidence intervals
confint(fit_d)

```
## Part D

```{r Q1D}
## Calculates raw residuals
residd= resid(fit_d)

print(residd)

```


Based off the values from part a, we know the y-intercept and the slope of the regression line. This give us the equation:

$$ \hat{y} = 10.137 - 0.0372x$$

## Part E

```{r Q1E}

##Print the error values
print(summaryd[6])

```

Based off the summary statistics generated in part b, we know that $\hat{\sigma}^2 = .3624$

## Part F

```{r Q1F}
## Predicts the 95% confidence interval based on the new rice of x = 100
mu0conf = predict(fit_d, newdata = data.frame(x = 100), interval = "confidence")

print(mu0conf)
```

## Part G
```{r Q1G}
## Predicts the 95% prediction interval based on the new rice of x = 100
mu0pred = predict(fit_d, newdata = data.frame(x = 100), interval = "prediction")

print(mu0pred)
```

Compared to the confidence interval (range = 0.647), the prediction interval (range = 1.8879) is much wider

## Part H
```{r}
## Pulls the r^2 value from summaryd
d_r_squared = summaryd['r.squared']
```

The R-squared value represents how much variance that can be explained by the independent variable, in this case plant height. 

# Question 2

## Part A
```{r}
library(tidyverse, warn.conflicts = FALSE)
## Data for question 2
x2 = c(1, 2, 3, 4, 5, 6, 7, 8, 9)
y2 = c(-2.08, -0.72, 0.28, 0.92, 1.20, 1.12, 0.68, -0.12, -1.28)

# Combines data into datafram
d2 = data.frame(x2,y2)

## Plots data
ggplot(d2, aes(x = x2, y = y2)) +
  geom_point()
```

## Part B

